---
title: "Illustration on Big5 data"
format: html
editor: visual
---

The purpose of the reported analyses here below is mainly to show the advantages of TSPCA over alternative approaches with equal weights and loadings based on sparse PCA methods with either sparse weights or sparse loadings. After all, a straightforward solution may seem to use the sparse loadings obtained from a sparse PCA analysis with sparseness on the loadings as the weights and the other way around. The advantages of TSPCA over such ad hoc approaches are as follows:

-TSPCA will have a higher proportion of explained variance than methods using a sparse loading approach (and with the weights put equal to the loadings and both properly scaled for fair comparison to TSPCA). This is, TSPCA optimizes fit for a solution with weights equal to loadings.

-TSPCA will have much more stable weights than methods using a sparse weights approach (and a higher proportion of explained variance when the loadings are set equal to the obtained weights and rescaling them properly).

An illustrative real data example is used to underpin our statements. First, we focus on the proportion of explained variance and second, on the stability of the zero/nonzero status of the weights using resampling.

# Data Preparation

## Imports

```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
rm(list = ls())
```

```{r}
library("qgraph") # Big5 data are part of qgraph package
```

## Big5 data

The data represent the scores of 500 respondents of the 240 items of the neo-pi-r (Costa & McCrae, 1992). Here, we use the data collected by Dolan et al. (2009) on 500 first year psychology students and available as part of the qgraph package. The neo-pi-r is a questionnaire that has been developed to measure the 5 OCEAN personality traits, using 48 items per trait.

```{r}
# obtain data
data("big5",package = "qgraph")
# Mean centering + scaling to unit variance
big5_std<-scale(big5,center = T,scale = T)

# Reorder the variables to OCEAN structure
itemlabels <- colnames(big5_std)
Oind <- grep("O",itemlabels)
Cind <- grep("C",itemlabels)
Eind <- grep("E",itemlabels)
Aind <- grep("A",itemlabels)
Nind <- grep("N",itemlabels)

OceanData <- big5_std[,c(Oind,Cind,Eind,Aind,Nind)]

# Obtain basic information
N <- dim(OceanData)[1]
J <- dim(OceanData)[2]
ssx <- sum(OceanData^2)

# Fix number of factors to the 5 personality traits
R <- 5 # 2 groups

```

# Regular PCA with W = P

First we perform regular / svd based PCA that automatically yields W = P (both orthogonal)

```{r}
svdX <- svd(OceanData, nu = R, nv = R) 
Loadings <- svdX$v 
Scores <- OceanData %*% Loadings 
PVEpca <- sum(svdX$d[1:R]^2)/ssx  
round(Loadings, 3) 
PVEpca
```

This solution has maximal PVE but does not show simple structure.

# EFA/PCA with sparse loadings

First, we conduct a sparse factor analysis with sparseness of the loadings. The seafar package (Le et al., 2026) is used that offers an implementation of the USLPCA method (Adachi & Trendafilov, 2016) that allows to impose the number of (non)zero loadings. Different from USLPCA, the factors can be correlated. The resulting factor scores are standardized.

## Imports

```{r}
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
if (!requireNamespace("PCAtools", quietly = TRUE)) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
  BiocManager::install("PCAtools")
}
# EFA.dimensions (another seafar dependency) was archived
if (!requireNamespace("EFA.dimensions", quietly = TRUE)) {
  efa_deps <- c("psych", "polycor", "EFAtools", "mirt", "GPArotation", "lavaan", "semTools")
  missing_deps <- efa_deps[!sapply(efa_deps, requireNamespace, quietly = TRUE)]
  if (length(missing_deps) > 0) install.packages(missing_deps)
  remotes::install_version("EFA.dimensions", version = "0.1.8.8", repos = "https://cran.r-project.org", upgrade = "never")
}
if (!requireNamespace("seafar", quietly = TRUE)) {
  remotes::install_github("trale97/seafar@*release")
}
```

```{r}
set.seed(1808)
```

```{r}
library(seafar)  
nrnonzero <- 240 #Following the design of the questionnaire 
out_seafa <- seafar:::seafar_multistart(OceanData, nfactors = R, C = nrnonzero, INIT = "rational") 
Xhat <- out_seafa$scores%*%t(out_seafa$loadings) 
PVEseafa <- 1-(sum((OceanData-Xhat)^2))/ssx 
#Check of design: Do items designed for a trait load on same factor?
L <- 1
U <- L+47
for (i in 1:5){
  print(colSums(out_seafa$loadings[L:U,]!=0))
  L <- U+1
  U <- L+47
} 
PVEseafa
```

The obtained solution is sparse and has -upto rounding - same PVE as the PCA solution (which has the highest PVE by definition).

As an alternative to TSPCA, one might consider using the obtained loadings as the weights when calculating the factor scores (with proper scaling so that the scores have variance equal to one as assumed in the seafar model). To legitimize the use of TSPCA, the proportion of variance explained by this solution should be lower.

```{r}
# Calculate factor scores with seafar loadings 
T_seafa_unscaled <- OceanData %*% out_seafa$loadings 
# Rescaling 
T_norm <- colSums(T_seafa_unscaled^2)
f_diag <- diag(sqrt(N)/ sqrt(T_norm))
PW <- out_seafa$loadings%*%sqrt(f_diag)
T_seafa_scaled <- T_seafa_unscaled %*% sqrt(f_diag)
P_seafa_scaled <- out_seafa$loadings %*% sqrt(f_diag)
X_hatW <- T_seafa_scaled %*% t(P_seafa_scaled)  
X_hatW <- (OceanData%*%PW)%*%t(PW)
# Compare Xhat based on heuristic with Xhat seafar 
round(head(cbind(X_hatW[,1:3], Xhat[,1:3])),3)  
#Proportion of variance explained 
1-sum((OceanData-X_hatW)^2)/ssx 
```

# PCA with sparse weights

PCA with sparse weights was first developed by Zou et al. (2006) and was implemented in the elasticnet package. Another approach to PCA with sparse weights was proposed by Journée et al. (2010) and found to be a strong performer (Guerra-Urzola et al., 2023).

## Imports

```{r}
library(elasticnet)
```

## Set seed

```{r}
set.seed(510)
```

```{r}
out_spca <- elasticnet::spca(OceanData, K=R, para = rep(48,5), sparse="varnum") 
L <- 1
U <- L+47
for (i in 1:5){
  print(colSums(out_spca$loadings[L:U,]!=0))
  L <- U+1
  U <- L+47
} 
out_spca$pev

#recalculate pev, derive loadings
alpha <- t(OceanData)%*%OceanData%*%out_spca$loadings
z <- svd(alpha)
P_spca <- (z$u) %*% t(z$v)
Xhat_spca <- (OceanData%*%out_spca$loadings)%*%t(P_spca)
1-(sum((OceanData-Xhat_spca)^2))/ssx

#use spca weights as loadings
Xhat_spca <- (OceanData%*%out_spca$loadings)%*%t(out_spca$loadings)
1-(sum((OceanData-Xhat_spca)^2))/ssx

#gpower to obtain sparse weights
source("../Scripts/GPower_Functions.R")
out_gpower <- gpower_block_cardinality(X=OceanData, R, phi=5*48, mu = 1,cardinality_type = "total", center = TRUE, scale = FALSE,max_iter = 1000, tol = 1e-4, verbose = FALSE) 
  
Xhat_gpower <- (OceanData%*%out_gpower$weights)%*%t(out_gpower$loadings)
1-(sum((OceanData-Xhat_gpower)^2))/ssx
L <- 1
U <- L+47
for (i in 1:5){
  print(colSums(out_gpower$weights[L:U,]!=0))
  L <- U+1
  U <- L+47
} 

```

The PVE for spca is higher than for the approach with sparse loadings (as can be expected). Yet, stability of the weights is an issue, for spca as well as for gpower (see the resampling analysis below).

# W = P analysis

```{r}
source("../Scripts/TSPCA_Functions.R")
epsilon <- 1e-8
phi <- nrnonzero #number of nonzero loadings, per component: 48 items per trait
###set penalty to balance fit with penalty
rho <- sum(OceanData^2)/R #||X-XWP'||^2 is max ||X||^2 while ||W-P||^2 is max R
# Equality-constrained approach (TSPCA): loadings forced exactly equal to weights.
# cardinality_type = "per_component" guarantees exactly `phi` (48) nonzero weights
# in EACH of the 5 components -- the global/matrix-wide budget used elsewhere in
# this codebase gives no such guarantee and could in principle starve a trait of
# nonzero weights entirely (which would abort with "zero column in W").
out_tspca <- NULL; best_loss <- Inf
INIT <- PW
for (m in 1:20) {
  set.seed(100 + m)
  res <- TSPCA(OceanData, INIT, R, 1e-8, phi=48, rho, constrained=1, MaxIter=200,  cardinality_type = "per_component")
  if (res$Residual < best_loss)
  { best_loss <- res$Residual; out_tspca <- res }
}

PVEtspca <- 1-out_tspca$Residual
#round(cbind(X,xhat),4)
round(cbind(out_tspca$weights[,1:2],out_tspca$loadings[,1:2]),3)
PVEtspca
Xhat_tspca <- (OceanData%*%out_tspca$weights)%*%t(out_tspca$loadings)
1-(sum((OceanData-Xhat_tspca)^2))/ssx

#Check of design, as done above for seafar and spca: do items designed for a
#trait load on the TSPCA component for that same trait? Since weights = loadings
#here, checking one checks both.
L <- 1
U <- L+47
for (i in 1:5){
  print(colSums(out_tspca$weights[L:U,]!=0))
  L <- U+1
  U <- L+47
}
```

#Instability of weights, not of loadings

Use Hetvi's idea to create slightly perturbed data matrices.

```{r}
nsize <- 200 #resampling
nrreplic <- 10
seafarLoadings <- c()
spcaweights <- c()
gpowerweights <- c()
tspcaweights <- c()
spcazero <- matrix(0, nrow = J, ncol = R)
seafarzero <- matrix(0, nrow = J, ncol = R)
gpowerzero <- matrix(0, nrow = J, ncol = R)
tspcazero <- matrix(0, nrow = J, ncol = R)

#creation of reference pattern
RefPattern <- matrix(0, nrow = J, ncol = R)
L <- 1
for (r in 1:R) {
  U <- L + 47
  RefPattern[L:U, r] <- 1
  L <- U + 1
}

for (replic in 1:nrreplic){   # Noise
  index <- sample(500,nsize)
  SampleData <- OceanData[index,]
  ssx <- sum(SampleData^2)
  #spca: instable loadings when repeated
  out_spca <- elasticnet::spca(SampleData, K=R, para = rep(48,R), sparse="varnum")
  spca_aligned <- align_to_zero_pattern(out_spca$loadings, RefPattern)$aligned
  spcaweights <- cbind(spcaweights, spca_aligned)
  spcazero <- spcazero+(spca_aligned==0)
  out_gpower <- gpower_block_cardinality(X=SampleData, R, phi=5*48, mu = 1,cardinality_type = "total", center = TRUE, scale = FALSE,max_iter = 1000, tol = 1e-4, verbose = FALSE)
  gpower_aligned <- align_to_zero_pattern(out_gpower$weights, RefPattern)$aligned
  gpowerweights <- cbind(gpowerweights, gpower_aligned)
  gpowerzero <- gpowerzero+(gpower_aligned==0)
  out_seafa <- seafar:::seafar_multistart(SampleData, nfactors = R, C = nrnonzero, initloadings = PW)
  Xhat <- out_seafa$scores%*%t(out_seafa$loadings)
  seafar_aligned <- align_to_zero_pattern(out_seafa$loadings, RefPattern)$aligned
  seafarLoadings <- cbind(seafarLoadings, seafar_aligned)  
  seafarzero <- seafarzero+(seafar_aligned==0) 
  out_tspca <- TSPCA(SampleData, PW, R, epsilon, phi=5*48,rho, constrained = 1, MaxIter=200,cardinality_type = "total")
  tspca_aligned <- align_to_zero_pattern(out_tspca$weights, RefPattern)$aligned
  tspcaweights <- cbind(tspcaweights, tspca_aligned)
  tspcazero <- tspcazero+(tspca_aligned==0)
}

```

Here below we calculate, for each method, the percentage of the 1200 weights (loadings for seafar) that had the same zero/nonzero status in all resamples. TSPCA yields very stable results (98.5%; approximately 960 weights are zero and 240 are nonzero in each resample), followed by the sparse loadings approach seafar (89.9%). The two sparse weights methods are much less stable: gpower (60.3%) and spca (55.5%) show consistent zero/nonzero status for little more than half of the weights.

```{r}
# Per-component variance across resamples (not pooled across components)
spca_var_by_comp <- sapply(1:R, function(r) apply(spcaweights[, seq(r, R*nrreplic, by = R)], 1, var))
tspca_var_by_comp   <- sapply(1:R, function(r) apply(tspcaweights[,   seq(r, R*nrreplic, by = R)], 1, var))
colnames(spca_var_by_comp) <- colnames(tspca_var_by_comp) <- paste0("Comp", 1:R)
cat("Mean per-item weight variance across resamples, by component:\n")
cat("spca: "); print(round(colMeans(spca_var_by_comp), 4))
cat("TSPCA: "); print(round(colMeans(tspca_var_by_comp), 4))

###COMPARE ON ZERO/NONZERO STATUS (scale-free: does each entry keep the same
### zero/nonzero status across all nrreplic resamples?)
sum(tspcazero==0)
sum(tspcazero==nrreplic)

sum(spcazero==0)
sum(spcazero==nrreplic)

# Same information as a single stability rate per method, easier to report directly:
cat(sprintf("TSPCA: %.1f%% of weights had unanimous zero/nonzero status across resamples\n",
            100 * (sum(tspcazero==0) + sum(tspcazero==nrreplic)) / length(tspcazero)))
cat(sprintf("spca:  %.1f%% of weights had unanimous zero/nonzero status across resamples\n",
            100 * (sum(spcazero==0) + sum(spcazero==nrreplic)) / length(spcazero)))
cat(sprintf("seafar:  %.1f%% of loadings had unanimous zero/nonzero status across resamples\n",
            100 * (sum(seafarzero==0) + sum(seafarzero==nrreplic)) / length(seafarzero)))
cat(sprintf("gpower:  %.1f%% of weights had unanimous zero/nonzero status across resamples\n",
            100 * (sum(gpowerzero==0) + sum(gpowerzero==nrreplic)) / length(gpowerzero)))
```
